---
title: "IRS Normalisation approach CSF data"
author: "JJ"
date: "09/10/2024"
output:
  html_document: default
  pdf_document: default
---

```{r, include = FALSE}

library(dplyr)
library(tidyr)
library(ggplot2)
library(readxl)
library(ggpubr)
library(survminer)
library(survival)
library(tableone)
library(tidyverse)
library(janitor)
library(limma) #for plotDensities function
library(ggfortify)
library(corrplot)
library(gmm)
library(pals)
library(corrplot)
library(circlize)
library(randomcoloR)
library(ggVennDiagram)
library(psych)
library(data.table)

#rm(list = ls())
```


```{r }
## Pre- normalization data processing 

#load in and clean data
#load in data


#für Windows:
#Prot_raw = read.delim("C:/Users/AG_Oellerich/OneDrive - Johann Wolfgang Goethe Universität/Frankfurt_Oellerich/2024-08_CSF_Klinsing_cohort2/MS results/proteinGroups.txt", stringsAsFactors = FALSE, colClasses = "character")
Prot_raw = read.delim("/Users/josefinejakob/Library/CloudStorage/OneDrive-JohannWolfgangGoetheUniversität/Frankfurt_Oellerich/2024-08_CSF_Klinsing_cohort2/MS results/proteinGroups.txt", stringsAsFactors = FALSE, colClasses = "character")

## Pre- normalization data processing 

#To include only valid protein information, proteins that were identified only by site, reverse or with potential contaminant were excluded. 


#clean column names for spaces 
#names(Prot_raw) <- str_replace_all(names(Prot_raw), c(" " = "_"))
Prot_raw <- as.data.frame(Prot_raw)
```

#Analysis WITHOUT contaminants
```{r }
#remove "remove only by side" identified proteins 
Prot_raw1 <- dplyr::filter(Prot_raw, Potential.contaminant != "+") %>%
  dplyr::filter(Reverse != "+") %>%
  dplyr::filter(Only.identified.by.site != "+")

#Pull out relevant information of Reporter.intensity.corrected
Prot_raw1 <- Prot_raw1 %>%
  dplyr::select(Protein.IDs, contains("Reporter.intensity.corrected")) %>%
  filter(!str_detect(Protein.IDs, "^REV"))
#convert Zero values to NA
Prot_raw1[,-1][Prot_raw1[,-1] == "0"] <- NA
#remove rows with only NA 
Prot_raw2 <- Prot_raw1 %>%
  filter(rowSums(is.na(Prot_raw1[,-1])) != ncol(Prot_raw1[,-1]))
#TRANFORM INTENSITY COLUMNS AS NUMERIC
intensity.names = grep("^Reporter.intensity.corrected", names(Prot_raw2), value = TRUE)
Prot_raw2[intensity.names] = sapply(Prot_raw2[intensity.names], as.numeric)
glimpse(Prot_raw2)

#remove redundant colname information 
col_headers <- colnames(Prot_raw2)
col_headers <- str_replace(col_headers, "Reporter.intensity.corrected", "")
colnames(Prot_raw2) <- col_headers

glimpse(Prot_raw2)

#rename_IDs <- read_excel("C:/Users/AG_Oellerich/OneDrive - Johann Wolfgang Goethe Universität/Frankfurt_Oellerich/2024-08_CSF_Klinsing_cohort2/MS results/rename_sample_IDs.xlsx")
rename_IDs <- read_excel("/Users/josefinejakob/Library/CloudStorage/OneDrive-JohannWolfgangGoetheUniversität/Frankfurt_Oellerich/2024-08_CSF_Klinsing_cohort2/MS results/rename_sample_IDs.xlsx")

# Create new column names by appending the condition
new_col_names <- paste(rename_IDs$MS_ID, rename_IDs$Group, rename_IDs$Sample_ID, sep = "_")  #The paste() function appends the Group to the column names using an underscore (_).

# Ensure the length of new_col_names matches the number of columns being renamed
if(length(new_col_names) == length(2:127)) {
  # Assign the new column names directly to the specified columns of Prot_raw2
  colnames(Prot_raw2)[2:127] <- new_col_names
} else {
  stop("The number of new column names doesn't match the number of columns to be renamed.")
}

# Check the updated df1
glimpse(Prot_raw2)
```




### Pre-normalization data inspection 
```{r }
#Tabular expression of data with uniprot ids excluded 

Prot_raw_init <- Prot_raw2 %>%
  dplyr::filter(!grepl("^REV", Protein.IDs))

# Raw data column totals

format(round(colSums(Prot_raw2[,-1], na.rm=TRUE), digits = 0), big.mark = ",")
```

```{r }
#Raw data Boxplot

boxplot(log2(Prot_raw_init[, -1]),
   notch = TRUE,
  main = 'RAW data without normalization',
  xlab = 'TMT Samples',
  ylab = 'log2 of Intensity'
)

```
#Raw data in density plots (like a distribution histogram)

```{r, echo = FALSE}
# can also look at density plots (like a distribution histogram)
plotDensities(log2(Prot_raw_init[, -1]),
              main = 'Raw data',
              legend = FALSE)
```

```{r, fig.width= 15}

#Total number of proteins quantified per sample

# Calculate the number of non-NA values per column
non_na_counts <- Prot_raw_init %>%
  select(-Protein.IDs) %>% # Exclude the Protein.IDs column
  summarise(across(everything(), ~ sum(!is.na(.))))%>%
  pivot_longer(everything(), names_to = "Sample", values_to = "Non_NA_Counts")

# Create the bar diagram
Proteins_per_sample_plot <- ggplot(non_na_counts, aes(x = Sample, y = Non_NA_Counts)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  theme(axis.text.x = element_text(angle=90, hjust=1, vjust=0.5))+
  labs(title = "Number of Proteins Identified by Sample",
       x = "Sample",
       y = "Number of Non-NA Values") 
  
#ggsave("C:/Users/AG_Oellerich/OneDrive - Johann Wolfgang Goethe Universität/Frankfurt_Oellerich/2024-08_CSF_Klinsing_cohort2/MS results/plots/2024-10-10_proteins_per_sample.png", Proteins_per_sample_plot, width = 350, height = 200, units = "mm")

```

```{r }

#create PCA df
dataPCA_raw <- Prot_raw_init[,-1] %>%
  drop_na() %>%
  t() %>%
  as.data.frame()
##make rownames to first column
dataPCA_raw <- dataPCA_raw %>%
  as.data.frame() %>%
  tibble::rownames_to_column("Sample_ID") 

data_PCA_ov_raw <- dataPCA_raw %>%
  separate(Sample_ID, c("Label", "Replicate", "Plex", "Condition", "SampleID"), sep = "_")

rownames(data_PCA_ov_raw) <- dataPCA_raw$Sample_ID

#create second dummy ds
dataPCA_raw1 <- dataPCA_raw[,-1]
#remove first column
dataPCA_raw$Sample_ID<- NULL 
#create PCA object prcomp
data_pca_raw <- prcomp(dataPCA_raw1, center = TRUE, scale. = TRUE)
#plot PCA object with cell line information

n <- 6

palette <- distinctColorPalette(n)

autoplot(data_pca_raw, data = data_PCA_ov_raw, label= TRUE, colour = "Plex", legend = FALSE) + 
  theme_classic() +
  scale_color_manual(values = palette) +
  guides(color = "none")

```

```{r }

#Missingness rate

# Calculate missingness rate per protein
missingness_per_protein <- Prot_raw_init %>%
  # Gather all sample columns into long format
  pivot_longer(cols = -Protein.IDs, names_to = "Sample", values_to = "Intensity") %>%
  group_by(Protein.IDs) %>%
  # Calculate number of non-missing values per protein
  summarize(total_values = n(),
            non_missing_values = sum(!is.na(Intensity)))%>%
  # Calculate percentage of missing values per protein
  mutate(missingness_rate = (100-((non_missing_values / total_values)) * 100))

missingness_plot <- missingness_per_protein %>% 
  ggplot(aes(x=missingness_rate)) +
  geom_histogram(col = "black", bins = 30) +
  cowplot::theme_cowplot() +
  ggtitle("histogram of protein missingness, all samples")+
 labs(x="Missingness rate (%)", 
      y="Protein Count")


#ggsave("C:/Users/AG_Oellerich/OneDrive - Johann Wolfgang Goethe Universität/Frankfurt_Oellerich/2024-08_CSF_Klinsing_cohort2/MS results/plots/2024-10-10_missingness.png", missingness_plot, width = 250, height = 200, units = "mm")


```

#Sample loading (Sl) normalization
```{r }
  #We should have the total signals in each channel summing to the same value. We have significant differences that we can correct with some basic normalizing. We can average the numbers below and compute normalization factors to make the sums end up the same. -> corrects for small sample loading and labeling reaction efficiency differences.
#summary: Summe jedes Channels/Probe -> mean aller Proben (mittlere Spaltengesamtintensität) -> Spaltenkorrekturfaktor (mittlere Spaltengesamtintenstität )/jeweilige Spaltenintensität --> Anwendung des Faktors auf jede Spalte
#Code von pwilmart.github

# separate the TMT data by experiment
# we do not need to do this for the normalization factor calculation here,
# but we will need these data frames for the IRS step below.
exp1_raw <- Prot_raw_init %>% select("Protein.IDs", contains("_825"))

exp2_raw <- Prot_raw_init %>% select("Protein.IDs", contains("_826"))

exp3_raw <- Prot_raw_init %>% select("Protein.IDs", contains("_827"))

exp4_raw <- Prot_raw_init %>% select("Protein.IDs", contains("_828"))

exp5_raw <- Prot_raw_init %>% select("Protein.IDs", contains("_829"))

exp6_raw <- Prot_raw_init %>% select("Protein.IDs", contains("_830"))


# first basic normalization is to adjust each TMT experiment to equal signal per channel
# figure out the global scaling value
target <- mean(c(colSums(exp1_raw[,-1], na.rm = TRUE), colSums(exp2_raw[,-1], na.rm=TRUE), colSums(exp3_raw[,-1], na.rm=TRUE), colSums(exp4_raw[,-1], na.rm=TRUE),colSums(exp5_raw[,-1], na.rm=TRUE),colSums(exp6_raw[,-1], na.rm=TRUE) ))

# do the sample loading normalization before the IRS normalization
# there is a different correction factor for each column
norm_facs <- target / colSums(exp1_raw[,-1], na.rm = TRUE)
exp1_sl <- sweep(exp1_raw[-1], 2, norm_facs, FUN = "*")

norm_facs2 <- target / colSums(exp2_raw[,-1], na.rm = TRUE)
exp2_sl <- sweep(exp2_raw[,-1], 2, norm_facs2, FUN = "*")

norm_facs3 <- target / colSums(exp3_raw[,-1], na.rm = TRUE)
exp3_sl <- sweep(exp3_raw[,-1], 2, norm_facs3, FUN = "*")

norm_facs4 <- target / colSums(exp4_raw[,-1], na.rm = TRUE)
exp4_sl <- sweep(exp4_raw[,-1], 2, norm_facs4, FUN = "*")

norm_facs5 <- target / colSums(exp5_raw[,-1], na.rm = TRUE)
exp5_sl <- sweep(exp5_raw[,-1], 2, norm_facs5, FUN = "*")


norm_facs6 <- target / colSums(exp6_raw[,-1], na.rm = TRUE)
exp6_sl <- sweep(exp6_raw[,-1], 2, norm_facs6, FUN = "*")




# make a pre-IRS data frame after sample loading normalizations
data_sl <- cbind(exp1_raw%>%select("Protein.IDs"), exp1_sl, exp2_sl, exp3_sl, exp4_sl, exp5_sl, exp6_sl)

boxplot(log2(data_sl[,-1]), 
        notch = TRUE, main = "Sample Loading (SL) normalized data",
        xlab = 'TMT Sample', ylab = 'log2 of Intensity')
```

```{r, fig.width=15 }

plotDensities(log2(data_sl[,-1]), main = "SL normalization")

```

```{r }

#create PCA df
dataPCA_sl <- data_sl[,-1] %>%
  drop_na() %>% #drops rows with NA values completely
  t() %>% #transforms rows <-> columns
  as.data.frame() 
##make rownames to first column
dataPCA_sl <- dataPCA_sl %>%
  as.data.frame() %>%
  tibble::rownames_to_column("Sample_ID") 

data_PCA_ov_sl <- dataPCA_sl %>%
  separate(Sample_ID, c("Label", "Replicate", "Plex", "Condition", "SampleID"), sep = "_")

rownames(data_PCA_ov_sl) <- dataPCA_sl$Sample_ID

#create second dummy ds
dataPCA_sl1 <- dataPCA_sl
#remove first column
dataPCA_sl1$Sample_ID<- NULL 
#create PCA object prcomp
data_pca_sl <- prcomp(dataPCA_sl1, center = TRUE, scale. = TRUE)
#plot PCA object 

n <- 6
palette <- distinctColorPalette(n)

autoplot(data_pca_sl, data = data_PCA_ov_sl, label= TRUE, colour = "Plex", legend = FALSE) + 
  theme_classic() +
  scale_color_manual(values = palette) +
  guides(color = "none")

```

```{r }
# check column totals
format(round(colSums(data_sl[,-1], na.rm=TRUE), digits = 0), big.mark = ",")
```

## Internal reference scaling (IRS) normalization between plexes 
```{r }
#make a dataframe of the reference channels per plex 
irs_factors <- data_sl %>% 
  dplyr::select(contains("Std"))

#calculate the geometric mean per row of the reference channels 
irs_factors$geomean <- apply(irs_factors, 1, function(x) exp(mean(log(x), na.rm = TRUE)))

#pull out sample names to identify
irs_factor_vec = as.vector(names(irs_factors))

#create the protein and reference channel wise scaling factor 
for(i in irs_factor_vec) {
  factor <-  irs_factors$geomean / (irs_factors %>% dplyr::select(i))
  irs_factors[,paste0("fac_",i)] <- factor
}

#step-by-step breakdown of the code:
  #for(i in irs_factor_vec) {: This line initiates a for loop that iterates over each element in the irs_factor_vec.
  #factor <- irs_factors$geomean / (irs_factors %>% dplyr::select(i)):
  #irs_factors$geomean: Accesses the geomean column in the irs_factors data frame.
  #irs_factors %>% dplyr::select(i): Uses dplyr::select to select the column named by the current value of i in the irs_factors data frame.
  #The division irs_factors$geomean / (irs_factors %>% dplyr::select(i)) calculates the factor.
  #irs_factors[,paste0("fac_",i)] <- factor: This line creates a new column in the irs_factors data frame. The name of the new column is constructed by concatenating "fac_" with the current value of i. The calculated      #factor is assigned to this new column.


data_irs_825_R1 <- cbind(data_sl%>%select("Protein.IDs"), data_sl%>% select(contains("R1_825")) * irs_factors$fac_.1_R1_825_Std_Std)
data_irs_825_R2 <- cbind(data_sl%>%select("Protein.IDs"), data_sl%>% select(contains("R2_825")) * irs_factors$fac_.1_R2_825_Std_Std)

data_irs_826_R1 <- cbind(data_sl%>%select("Protein.IDs"), data_sl%>% select(contains("R1_826")) * irs_factors$fac_.1_R1_826_Std_Std)
data_irs_826_R2 <- cbind(data_sl%>%select("Protein.IDs"), data_sl%>% select(contains("R2_826")) * irs_factors$fac_.1_R2_826_Std_Std)

data_irs_827_R1 <- cbind(data_sl%>%select("Protein.IDs"), data_sl%>% select(contains("R1_827")) * irs_factors$fac_.1_R1_827_Std_Std)
data_irs_827_R2 <- cbind(data_sl%>%select("Protein.IDs"), data_sl%>% select(contains("R2_827")) * irs_factors$fac_.1_R2_827_Std_Std)

data_irs_828_R1 <- cbind(data_sl%>%select("Protein.IDs"), data_sl%>% select(contains("R1_828")) * irs_factors$fac_.1_R1_828_Std_Std)
data_irs_828_R2 <- cbind(data_sl%>%select("Protein.IDs"), data_sl%>% select(contains("R2_828")) * irs_factors$fac_.1_R2_828_Std_Std)

data_irs_829_R1 <- cbind(data_sl%>%select("Protein.IDs"), data_sl%>% select(contains("R1_829")) * irs_factors$fac_.1_R1_829_Std_Std)
data_irs_829_R2 <- cbind(data_sl%>%select("Protein.IDs"), data_sl%>% select(contains("R2_829")) * irs_factors$fac_.1_R2_829_Std_Std)

data_irs_830_R1 <- cbind(data_sl%>%select("Protein.IDs"), data_sl%>% select(contains("R1_830")) * irs_factors$fac_.1_R1_830_Std_Std)
data_irs_830_R2 <- cbind(data_sl%>%select("Protein.IDs"), data_sl%>% select(contains("R2_830")) * irs_factors$fac_.1_R2_830_Std_Std)

data_irs <- data_irs_825_R1 %>%
  left_join(data_irs_825_R2, by = "Protein.IDs") %>%
  left_join(data_irs_826_R1, by = "Protein.IDs") %>%
  left_join(data_irs_826_R2, by = "Protein.IDs") %>%
  left_join(data_irs_827_R1, by = "Protein.IDs") %>%
  left_join(data_irs_827_R2, by = "Protein.IDs") %>%
  left_join(data_irs_828_R1, by = "Protein.IDs") %>%
  left_join(data_irs_828_R2, by = "Protein.IDs")%>%
  left_join(data_irs_829_R1, by = "Protein.IDs")%>%
  left_join(data_irs_829_R2, by = "Protein.IDs")%>%
  left_join(data_irs_830_R1, by = "Protein.IDs")%>%
  left_join(data_irs_830_R2, by = "Protein.IDs")

#library(openxlsx)
#excel file saved below with Protein Description columns etc.

```

```{r }

# see what the IRS data look like
boxplot(log2(data_irs[,-1]), 
        main = "Internal Reference Scaling (IRS) normalized data",
        xlab = 'TMT Sample', ylab = 'log2 of Intensity', notch = TRUE)

```

```{r , fig.width=15}


# can also look at density plots (like a distribution histogram)    
plotDensities(log2(data_irs[,-1]),  main = "IRS data")
```

```{r }

#create PCA df
dataPCA_irs <- data_irs[,-1] %>%
  drop_na() %>%
  t() %>%
  as.data.frame()
##make rownames to first column
dataPCA_irs <- dataPCA_irs %>%
  as.data.frame() %>%
  tibble::rownames_to_column("Sample_ID") 

data_PCA_ov_irs <- dataPCA_irs %>%
  separate(Sample_ID, c("Label", "Replicate", "Plex", "Condition", "SampleID"), sep = "_")

rownames(data_PCA_ov_irs) <- dataPCA_irs$Sample_ID

#create second dummy ds
dataPCA_irs1 <- dataPCA_irs
#remove first column
dataPCA_irs1$Sample_ID<- NULL 
#create PCA object prcomp
data_pca_irs <- prcomp(dataPCA_irs1, center = TRUE, scale. = TRUE)
#plot PCA object with cell line information

n <- 10
palette <- distinctColorPalette(n)

#library(ggrepel)

pca_irs_plot <- autoplot(data_pca_irs, data = data_PCA_ov_irs, label= TRUE, colour = "Condition", legend = FALSE) + 
  theme_classic() +
  scale_color_manual(values = palette) +
  guides(color = "none") 
#pca with z-score scaled data see below


#library(ggplot2)
#ggsave("C:/Users/AG_Oellerich/OneDrive - Johann Wolfgang Goethe Universität/Frankfurt_Oellerich/2024-08_CSF_Klinsing_cohort2/MS results/plots/2024-10-10_PCA_IRS.png", pca_irs_plot, width = 250, height = 200, units = "mm")

```

```{r }
# check column totals
format(round(colSums(data_irs[,-1], na.rm=TRUE), digits = 0), big.mark = ",")
```

```{r, fig.width=12}
# heatmap
library(pheatmap)
pheatmap_irs_plot <- pheatmap(data_irs[2:127], cluster_rows = FALSE, cluster_cols = TRUE, na_col = "white")




#ggsave("/Users/josefinejakob/Desktop/Frankfurt_Oellerich/2024-05_CSF_Klinsing/Protokolle-Results/MS results/plots/2024-06-24_heatmap_irs.png", pheatmap_irs_plot, width = 250, height = 200, units = "mm")

```



#Further processing (lg2, Imputation, Means of replicates)
```{r }
Prot_raw_txt <- Prot_raw %>% select(,c(1,2,6:8, 10:12, 56))
#DEFINE PROTEIN AND GENE NAMES
# Extract Protein name
Prot_raw_txt$Protein = sub("OS=.*","", sub(".*HUMAN","", Prot_raw_txt$Fasta.headers))
# Extract Gene ID
Prot_raw_txt$Gene <- sub(".*GN=","", sub("PE=.*", "", Prot_raw_txt$Fasta.headers))

data_irs_long <- data_irs %>% 
  left_join(Prot_raw_txt, by = "Protein.IDs") %>%
  left_join(missingness_per_protein[,c("Protein.IDs", "missingness_rate")], by= "Protein.IDs")

glimpse(data_irs_long)

library(openxlsx)
#write.xlsx(data_irs_long, "C:/Users/AG_Oellerich/OneDrive - Johann Wolfgang Goethe Universität/Frankfurt_Oellerich/2024-08_CSF_Klinsing_cohort2/MS results/2024-10-10_data_normalized_sl_irs.xlsx")

```

```{r, fig.width=10}
#data loading mac
#data_irs_long <- read.xlsx("/Users/josefinejakob/Library/CloudStorage/OneDrive-JohannWolfgangGoetheUniversität/Frankfurt_Oellerich/2024-08_CSF_Klinsing_cohort2/MS results/2024-10-10_data_normalized_sl_irs.xlsx")

###########Z-SCORE
# Function to perform Z-score normalization per row
z_normalize <- function(row) {
  row_mean <- mean(row, na.rm = TRUE)
  row_sd <- sd(row, na.rm = TRUE)
  (row - row_mean) / row_sd
}

# Apply the function to each row
data_irs_subset_scaled <- data_irs_long[,2:127]
glimpse(data_irs_subset_scaled)
data_irs_subset_scaled <- t(apply(data_irs_subset_scaled, 1, z_normalize)) #1: function should be applied over the rows.
                                                             #2: function should be applied over the columns.

# Replace the scaled subset back into the original dataframe
data_irs_scaled <- data_irs_long # Copy original dataframe to preserve unmodified rows

data_irs_scaled [, 2:127] <- data_irs_subset_scaled
glimpse(data_irs_scaled)


```

```{r }
#PCA with z-scaled data
#create PCA df

dataPCA_irs_scaled <- data_irs_scaled[,2:127] %>%
  drop_na() %>%
  t() %>%
  as.data.frame()
##make rownames to first column
dataPCA_irs_scaled <- dataPCA_irs_scaled %>%
  as.data.frame() %>%
  tibble::rownames_to_column("Sample_ID") 

data_PCA_scaled_ov_irs <- dataPCA_irs_scaled %>%
  separate(Sample_ID, c("Label", "Replicate", "Plex", "Condition", "SampleID"), sep = "_")

rownames(data_PCA_scaled_ov_irs) <- dataPCA_irs_scaled$Sample_ID

#create second dummy ds
dataPCA_irs_scaled1 <- dataPCA_irs_scaled
#remove first column
dataPCA_irs_scaled1$Sample_ID<- NULL 
#create PCA object prcomp
data_pca_irs_scaled <- prcomp(dataPCA_irs_scaled1, center = TRUE, scale. = TRUE)
#plot PCA object with cell line information

n <- 10
palette <- distinctColorPalette(n)

#library(ggrepel)

pca_irs_scaled_plot <- autoplot(data_pca_irs_scaled, data = data_PCA_scaled_ov_irs, label= TRUE, colour = "Condition", legend = FALSE) + 
  theme_classic() +
  scale_color_manual(values = palette) +
  guides(color = "none") 
#pca with z-score scaled data see below


#library(ggplot2)
#ggsave("C:/Users/AG_Oellerich/OneDrive - Johann Wolfgang Goethe Universität/Frankfurt_Oellerich/2024-06_Leber_Maus_Hornung/MS results/plots/2024-07-29_liver_Hornung_PCA_norm_scaled.png", pca_irs_scaled_plot, width = 400, height = 250, units = "mm")

```

#LOG2 Transformation
```{r }
library(openxlsx)


#windows:
data_irs_long <- read_excel("C:/Users/AG_Oellerich/OneDrive - Johann Wolfgang Goethe Universität/Frankfurt_Oellerich/2024-08_CSF_Klinsing_cohort2/MS results//2024-10-10_data_normalized_sl_ir.xlsx")

#mac
data_irs_long<- read_excel("/Users/josefinejakob/Library/CloudStorage/OneDrive-JohannWolfgangGoetheUniversität/Frankfurt_Oellerich/2024-08_CSF_Klinsing_cohort2/MS results/2024-10-10_data_normalized_sl_irs.xlsx")

#LOG2 Transformation
glimpse(data_irs_long)
data_lg2 <- data_irs_long %>% rename_at(vars(starts_with(".1_R1_825_Std_Std"):starts_with(".10_R2_830_Metastasis_E24-313")), ~ paste0("log2", .))


data_lg2[2:127] = log2(data_lg2[2:127])

glimpse(data_lg2)

#filter out proteins with missingness >70%
data_lg2f <- data_lg2 %>% filter(missingness_rate <= 70)



```
#Imputation with DreamAI
```{r install packages, echo=FALSE}
#Packages required prior to installing DreamAI

require("cluster")
require("survival")
require("randomForest")
require("missForest")
require("glmnet")
require("Rcpp")
require("foreach")
require("itertools")
require("iterators")
require("Matrix")
require("devtools")
if (!requireNamespace("BiocManager", quietly = TRUE))
install.packages("BiocManager")
  BiocManager::install(version = '3.20')
BiocManager::install("impute", version = "3.20")
require("impute")

```

```{r install DreamAI, echo=FALSE}
# install DreamAI requirements
require("remotes")
#install_github("WangLab-MSSM/DreamAI/Code")

devtools::install_github("WangLab-MSSM/DreamAI/Code", dependencies = TRUE)
```

```{r DreamAI example}
library(DreamAI)

imputed <- DreamAI(data_lg2f[,2:127],k=10,maxiter_MF = 10, ntree = 100,maxnodes = NULL,maxiter_ADMIN=30,tol=10^(-2),gamma_ADMIN=NA,gamma=50,CV=FALSE,fillmethod="row_mean",maxiter_RegImpute=10,conv_nrmse = 1e-6,iter_SpectroFM=40, method = c("KNN", "MissForest", "ADMIN", "Birnn", "SpectroFM", "RegImpute"),out="Ensemble")
data_lg2I <- imputed$Ensemble

#Now add the imputed columns 2:127 to the whole data frame (which includes gene names etc)
# Create a new data frame by replacing columns 2:127
data_lg2I_long <- data_lg2f
data_lg2I_long[, 2:127] <- data_lg2I

# Check structure of the new data frame
str(data_lg2I_long)

# Preview the first few rows
head(data_lg2I_long)

library(openxlsx)
write.xlsx(data_lg2I_long, "/Users/josefinejakob/Library/CloudStorage/OneDrive-JohannWolfgangGoetheUniversität/Frankfurt_Oellerich/2024-08_CSF_Klinsing_cohort2/MS results/2024-12-03_data_normalized_DreamAI_imputed.xlsx")
```

```{r, fig.height=15 }
#heatmap of DreamAI imputed data
library(pheatmap)

pheatmap_imputed_plot <- pheatmap(data_lg2I_long[2:127], cluster_rows = FALSE, cluster_cols = TRUE, na_col = "white")

#pheatmap of z-scaled data
# Function to perform Z-score normalization per row
z_normalize <- function(row) {
  row_mean <- mean(row, na.rm = TRUE)
  row_sd <- sd(row, na.rm = TRUE)
  (row - row_mean) / row_sd
}

# Apply the function to each row
data_lg2I_subset <- data_lg2I_long[,2:127]
glimpse(data_lg2I_subset)
data_lg2I_subset_scaled <- t(apply(data_lg2I_subset, 1, z_normalize)) #1: function should be applied over the rows.
                                                                                #2: function should be applied over the columns.

# Replace the scaled subset back into the original dataframe
data_lg2I_scaled <- data_lg2I_long # Copy original dataframe to preserve unmodified rows

data_lg2I_scaled [, 2:127] <- data_lg2I_subset_scaled
glimpse(data_lg2I_scaled)

pheatmap_imputed_scaled_plot <- pheatmap(data_lg2I_scaled[,2:127], cluster_rows = FALSE, cluster_cols = TRUE, na_col = "white")

```

#Mean of replicates
```{r }


#windows: 
data_lg2I_long <- read_excel("C:/Users/AG_Oellerich/OneDrive - Johann Wolfgang Goethe Universität/Frankfurt_Oellerich/2024-08_CSF_Klinsing_cohort2/MS results/2024-12-03_data_normalized_DreamAI_imputed.xlsx")

#show headers
names(data_lg2I_long)


calculate_replicate_means <- function(data_lg2I_long) {
 # Ensure input is a data frame
  data_lg2I_long <- as.data.frame(data_lg2I_long)
  
   # Extract column names
  column_names <- names(data_lg2I_long)
  
  # Filter out only the columns that start with "log2."
  log2_columns <- column_names[grepl("^log2\\.", column_names)]
  
  # Extract the unique pattern (Sample name with E24-xxxx) from the end of the column names
  patterns <- unique(gsub(".*_R[12]_.*?_(.*)", "\\1", log2_columns))  # Capture unique sample identifiers (E.g., Primary-CNS-tumor_E24-0864)
  
  print("Replicate Patterns:")
  print(patterns)
  
  for (pattern in patterns) {
    # Construct regex pattern for columns with R1 and R2 for each unique sample
    pattern_regex <- paste0("_R[12]_.*_", pattern, "$")
    
    # Find columns matching the current pattern (for both R1 and R2 replicates)
    replicate_cols <- grep(pattern_regex, log2_columns, value = TRUE, perl = TRUE)
    
    print(paste("Pattern:", pattern))
    print("Replicate Columns:")
    print(replicate_cols)
    
    if (length(replicate_cols) >= 2) {
      # If we find columns for both R1 and R2
      mean_col_name <- paste0("mean_", pattern)
      data_lg2I_long <- data_lg2I_long %>%
        mutate(!!mean_col_name := rowMeans(select(., all_of(replicate_cols)), na.rm = TRUE))
    }
  }
  
  return(data_lg2I_long)
}



# Apply the function to the data frame
data_lg2I_reps <- calculate_replicate_means(data_lg2I_long)

#write.xlsx(data_lg2I_reps, "C:/Users/AG_Oellerich/OneDrive - Johann Wolfgang Goethe Universität/Frankfurt_Oellerich/2024-08_CSF_Klinsing_cohort2/MS results/2024-12-04_data_DreamAI_imputed_means.xlsx")

```

```{r, fig.width=15 }
#z-scale the imputed_means data
glimpse(data_lg2I_reps)


# Function to perform Z-score normalization per row
z_normalize <- function(row) {
  row_mean <- mean(row, na.rm = TRUE)
  row_sd <- sd(row, na.rm = TRUE)
  (row - row_mean) / row_sd
}

# Apply the function to each row
data_lg2I_reps_subset <- data_lg2I_reps[,138:195]
glimpse(data_lg2I_reps_subset)
data_lg2I_subset_reps_scaled <- t(apply(data_lg2I_reps_subset, 1, z_normalize)) #1: function should be applied over the rows.
                                                                                #2: function should be applied over the columns.


# Replace the scaled subset back into the original dataframe
data_lg2I_reps_scaled <- data_lg2I_reps # Copy original dataframe to preserve unmodified rows

data_lg2I_reps_scaled [, 138:195] <- data_lg2I_subset_reps_scaled
glimpse(data_lg2I_reps_scaled)

pheatmap_means_scaled_plot <- pheatmap(data_lg2I_reps_scaled[,138:195], cluster_rows = FALSE, cluster_cols = TRUE, na_col = "white")
```

```{r, fig.height=15 }
#heatmap
library(pheatmap)

## To color samples according to the condition: Make rownames to first column and separate Sample_ID
data_heatmap_lg2I_reps <- data.frame(Sample_ID = colnames(data_lg2I_reps_scaled[,139:195]))

data_heatmap_ov_lg2I_reps <- data_heatmap_lg2I_reps %>%
  separate(Sample_ID, c("mean", "Condition", "SampleID"), sep = "_")

rownames(data_heatmap_ov_lg2I_reps) <- data_heatmap_lg2I_reps$Sample_ID

# Create annotation data frame
sample_annotations <- data_heatmap_ov_lg2I_reps %>%
  select(Condition)                 #select condition will remove mean and PLex (they would also be shown otherwise in the heatmap with colors)



# Define annotation colors
ann_colors <- list(
  Condition = c("Std" = "grey", "Metastasis" = "lightblue", "Primary-CNS-tumor" = "blue", "Lymphoma"="green", "Inflammatory" ="red", "Control" = "purple")
)




# Generate the heatmap with annotations


pheatmap_imputed_means <- pheatmap(data_lg2I_reps_scaled[,139:195],
         cluster_rows = TRUE,  # Enable clustering of rows
         cluster_cols = TRUE,  # Enable clustering of columns
         clustering_distance_rows = "euclidean", # Euclidean distance for rows
         clustering_distance_cols = "euclidean", # Euclidean distance for columns
         clustering_method = "complete", # Clustering method, e.g., complete linkage
         scale = "none", # Data is already scaled
         show_rownames = TRUE, # Display row names
         show_colnames = TRUE, # Display column names
         color = colorRampPalette(c("blue", "white", "red"))(100),
         annotation_colors = ann_colors,
         annotation_col = sample_annotations) # Color scheme


#Windows: 
#ggsave("C:/Users/AG_Oellerich/OneDrive - Johann Wolfgang Goethe Universität/Frankfurt_Oellerich/2024-08_CSF_Klinsing_cohort2/MS results/plots/2024-12-04_CSF_heatmap_DreamAI_imputed_means_.png", pheatmap_imputed_means, width = 300, height = 200, units = "mm")
```

```{r, echo = FALSE, fig.height=5, fig.width=10}

#PCA
#create PCA df
data_lg2I_reps_PCA <- data_lg2I_reps_scaled[,139:195] %>%
  drop_na() %>%
  t() %>%
  as.data.frame()

##make rownames to first column
dataPCA_lg2I_reps<- data_lg2I_reps_PCA %>%
  as.data.frame() %>%
  tibble::rownames_to_column("Sample_ID") 
  
data_PCA_ov_lg2I_reps <- dataPCA_lg2I_reps %>%
  separate(Sample_ID, c("mean", "Condition", "SampleID"), sep = "_")

rownames(data_PCA_ov_lg2I_reps) <- dataPCA_lg2I_reps$Sample_ID



#create second dummy ds
dataPCA_lg2I_reps_1 <- dataPCA_lg2I_reps
#remove first column
dataPCA_lg2I_reps_1$Sample_ID<- NULL 
#create PCA object prcomp
data_pca_lg2I_reps <- prcomp(dataPCA_lg2I_reps_1, center = TRUE, scale. = TRUE)
#plot PCA object with cell line information




#o <- 9
#palette <- distinctColorPalette(o)
palette <- c("Std" = "grey", "Metastasis" = "lightblue", "Primary-CNS-tumor" = "blue", "Lymphoma"="green", "Inflammatory" ="red", "Control" = "purple")

pca_lg2I_reps <- autoplot(data_pca_lg2I_reps, data = data_PCA_ov_lg2I_reps, label= FALSE, colour = "Condition", legend = FALSE) + 
  geom_text(aes(label = SampleID), vjust = 1.5, size = 4) +
  theme_classic() +
  scale_color_manual(values = palette) +
  guides(color = guide_legend(title = "Condition", size=4), size=4)+
  stat_ellipse(aes(color = Condition))

#Windows: 
#ggsave("C:/Users/AG_Oellerich/OneDrive - Johann Wolfgang Goethe Universität/Frankfurt_Oellerich/2024-08_CSF_Klinsing_cohort2/MS results/plots/2024-12-14_CSF_pca_DreamAI_imputed_means.png", pca_lg2I_reps, width = 350, height = 200, units = "mm")



# Count the number of samples per condition
condition_counts <- sample_annotations %>%
  group_by(Condition) %>%
  summarise(Count = n())

# View the counts
print(condition_counts)

```

## Group-wise analysis
```{r}

library(openxlsx)

#windows:
#data_lg2I_reps <- read_excel("C:/Users/AG_Oellerich/OneDrive - Johann Wolfgang Goethe Universität/Frankfurt_Oellerich/2024-08_CSF_Klinsing_cohort2/MS results/2024-10-10_data_imputed_means.xlsx")

data_lg2I_reps2 <- data_lg2I_reps[,c(1,130, 136, 139:195)]

count_inflammatory <- sum(grepl("Inflammatory", colnames(data_lg2I_reps2)))
rest_noInfl <- c("Control","Metastasis","Primary-CNS-tumor","Lymphoma")

count_rest_noInfl <- sum(colnames(data_lg2I_reps2) %>% str_detect(paste(rest_noInfl, collapse = "|")))


data_lg2I_reps_FC <- data_lg2I_reps2 %>% 
  mutate(FC_Infl_vs_Ctrl = apply(data_lg2I_reps2 %>% select(contains("Inflammatory"), contains("Control")), 1, function(x) {mean(x[1:16])- mean(x[17:20])})) %>% 
  mutate(
    # Calculate the mean intensity for Inflammatory group
    Mean_Inflammatory = rowMeans(select(., contains("Inflammatory")), na.rm = TRUE),
    
    # Calculate the mean intensity for the rest (Control, Metastasis, Primary-CNS-tumor, Lymphoma)
    Mean_Rest = rowMeans(select(., contains("Control"), contains("Metastasis"), 
                                contains("Primary-CNS-tumor"), contains("Lymphoma")), na.rm = TRUE),
    
    # Compute the fold change (Inflammatory vs Rest)
    FC_Infl_vs_Rest = Mean_Inflammatory - Mean_Rest
  )




``` 

##Foldchanges
```{r}
#Inflammatory vs Ctrl
library(tidyr)
library(rstatix)
library(ggrepel)

Infl_vs_Ctrl <- data_lg2I_reps2%>% 
  select(contains("Inflammatory"), contains("Control"), Protein.IDs) %>%
pivot_longer(contains("mean_")) %>%
  group_by(Protein.IDs) %>%
  mutate(Group = case_when(
    grepl("Inflammatory", name) ~ "Inflammatory",
    grepl("Control", name) ~ "Control",  FALSE ~ "Other" )) %>%
  t_test(value ~ Group)

vocano  <- data_lg2I_reps_FC %>% 
  left_join(Infl_vs_Ctrl [,c("p", "Protein.IDs")], by = "Protein.IDs") %>% 
  mutate(P_Infl_vs_Ctrl = -log10(p)) %>% 
  select(-p) 

Scatter_Infl_vs_Ctrl<- ggplot(vocano, mapping=aes(FC_Infl_vs_Ctrl , P_Infl_vs_Ctrl ))+
  geom_point()+
  ylab("-log(p)")+
  xlab("log2(FC)")+
  #expand_limits(x=c(-6,6), y=c(0, 6))+
  ggtitle("Infl_vs_Ctrl")+
  geom_point(data=vocano %>% filter(FC_Infl_vs_Ctrl<(-0.6) & P_Infl_vs_Ctrl >1.301), color="cadetblue3")+
  geom_point(data=vocano  %>% filter(FC_Infl_vs_Ctrl >(0.6) & P_Infl_vs_Ctrl >1.301), color="cadetblue3")+
geom_text_repel(data = vocano %>% 
                  filter((FC_Infl_vs_Ctrl > 0.6 | FC_Infl_vs_Ctrl  < -0.6) & P_Infl_vs_Ctrl> 1.301), 
                aes(label = Gene.names), color = "red") 
#geom_segment(aes(x = 1, y = 1.301, xend= 4, yend = 1.301), color="red", linewidth =1.2, linetype="dashed")+
 # geom_segment(aes(x = -4, y = 1.301, xend= -1, yend = 1.301), color="red", linewidth =1.2, linetype="dashed")+
 # geom_segment(aes(x = 1, y = 1.301, xend= 1, yend = 9), color="red", linewidth =1.2, linetype="dashed")+
 # geom_segment(aes(x = -1, y = 1.301, xend= -1, yend = 9), color="red", linewidth =1.2, linetype="dashed")

```

##Scatterplots
```{r}
#Inflammatory vs all others
library(tidyr)
library(rstatix)
library(ggrepel)

Infl_vs_rest <- data_lg2I_reps2 %>%
  select(contains("Inflammatory"), contains("Control"), contains("Metastasis"),
         contains("Primary-CNS-tumor"), contains("Lymphoma"), Protein.IDs) %>%
  pivot_longer(cols = -Protein.IDs, names_to = "Sample", values_to = "Value") %>%
  mutate(Group = case_when(
    grepl("Inflammatory", Sample) ~ "Inflammatory",
    grepl("Control|Metastasis|Primary-CNS-tumor|Lymphoma", Sample) ~ "Rest",
    TRUE ~ NA_character_)) %>%
  filter(!is.na(Group)) %>%  # Remove rows without a valid group
  group_by(Protein.IDs) %>%
  t_test(Value ~ Group)  # Perform t-test

vocano  <- data_lg2I_reps_FC %>% 
  left_join(Infl_vs_rest [,c("p", "Protein.IDs")], by = "Protein.IDs") %>% 
  mutate(P_Infl_vs_Rest = -log10(p)) %>% 
  select(-p) 

Scatter_Infl_vs_rest <- ggplot(vocano, mapping=aes( FC_Infl_vs_Rest , P_Infl_vs_Rest ))+
  geom_point()+
  ylab("-log(p)")+
  xlab("log2(FC)")+
  #expand_limits(x=c(-6,6), y=c(0, 6))+
  ggtitle("Infl_vs_Rest")+
  geom_point(data=vocano %>% filter( FC_Infl_vs_Rest<(-0.6) & P_Infl_vs_Rest >1.301), color="cadetblue3")+
  geom_point(data=vocano  %>% filter( FC_Infl_vs_Rest >(0.6) & P_Infl_vs_Rest >1.301), color="cadetblue3")+
geom_text_repel(data = vocano %>% 
                  filter(( FC_Infl_vs_Rest > 0.6 |  FC_Infl_vs_Rest  < -0.6) & P_Infl_vs_Rest> 1.301), 
                aes(label = Gene.names), color = "red") 
#geom_segment(aes(x = 1, y = 1.301, xend= 2, yend = 1.301), color="red", linewidth =1.2, linetype="dashed")+
 # geom_segment(aes(x = -2, y = 1.301, xend= -1, yend = 1.301), color="red", linewidth =1.2, linetype="dashed")+
  #geom_segment(aes(x = 1, y = 1.301, xend= 1, yend = 5), color="red", linewidth =1.2, linetype="dashed")+
 # geom_segment(aes(x = -1, y = 1.301, xend= -1, yend = 5), color="red", linewidth =1.2, linetype="dashed")

#ggsave("C:/Users/AG_Oellerich/OneDrive - Johann Wolfgang Goethe Universität/Frankfurt_Oellerich/2024-08_CSF_Klinsing_cohort2/MS results/plots/2024-12-14_CSF_Scatter_DreamAI_imputed_Inflammation.png", pca_lg2I_reps, width = 350, height = 200, units = "mm")

```
´



